---
title: "APCtools - Main Functionality"
author: "Alexander Bauer, Maximilian Weigert, Pauline Hohenemser"
date: "`r format(Sys.time(), '%d.%B %Y')`"
output:
  html_document:
    toc: yes
  vignette: >
    %\VignetteIndexEntry{Functionality of TouristClust package}
    %\usepackage[utf8]{inputenc}
    %\VignetteEngine{knitr::rmarkdown}
  rmarkdown::html_vignette: default
editor_options:
  chunk_output_type: console
---

This document gives an overview of the functionality provided by the R package
`APCtools`.

An Age-Period-Cohort analysis is used to disentangle trends of social/health behvaiors into three components:
Age effect (life cycle) – This represents the effect associated with chronological age.
Period effect (macro-level developments) – This represents the effect associated with time period that affects all age groups at the same time.
Cohort effect (generational membership) – Epidemiologists and social scientists are often interested in birth cohort, who experienced the same historical and social events at the same age. Cohort effect thus is often used to refered to the variation due to these unique experiences.
These three are linked through the identity age + cohort = period.

The package includes functions for descriptive plots, innovative visualization techniques for APC models like ridgeline matrix for metric and categorical variables, estimation of Partial effect plots.Furthermore smoothed and non-smoothed model based heatmaps, effect plots of linear and nonlinear effects, and Hexamaps for a better APC analysis. 

```{r packages, message = FALSE}
library(APCtools)
library(dplyr)
library(mgcv)

# set the global theme of all plots
library(ggplot2)
theme_set(theme_minimal())
```

### Data overview
APC analyses require long-term panel or repeated cross-sectional data (Yang and Land, 2013). In this example, we investigates how age, period, and birth cohorts are related to altering travel distances in Germany between 1971 and 2018. 

```{r data preparation}
data(travel)
data(drug_deaths)
```

# Descriptive analyses

### 1D visualization function for metric and categorical variables
Plot_variable will show you the distribution of your data.In the case of a metric variable, the function will 
give you the mean travel distance, 25th/75th percentile and the lowest/highest data point (excluding any outliers) for each year, to help illustrate the magnitude of the difference as well as the uncertainty over the years. Outliers are also visible as dots outside of the boxplot. If the difference in your values are big, it is possible to use a log2-scale to get a useful visualization. 

If your variable is categorical, bar charts are used to represent counts of the values. By default, the function will show 
you each category as a percentage  of the whole data in each year, to see the magnitude of each category over the periods. If you add "stack" the data will be shown in absolute numbers for each year and category to get a sense of the quantity of the data points over the years.

```{r, warning = FALSE}
# plot a metric variable
plot_variable(dat = travel, variable = "mainTrip_distance", apc_dimension = "period", log_scale = TRUE)

# plot a categorical variable
plot_variable(dat = travel, variable = "household_size", apc_dimension = "period")
plot_variable(dat = travel, variable = "household_size", apc_dimension = "period", geomBar_position = "stack")
```

### Density matrix for metric and categorical variables

You are able to define the "breaking points" in your data as you need them, as well as adding a label to each level. 
In this step, you should also define the age-, period-, and cohort-groups.

These ridgeline plots shows the density of your data on a log scale in a matrix structure, split up in the age/period 
groups you defined prior.

If you want to highlight your cohort-groups you can show them if you add "highlight_diagonals" and select the ones you 
want to be marked.

To further illustrate the plot it is possible to visualize your "breaking points" with different colours for the 
different categories.

You can also change the depiction if you change your temporal dimension from "age" to "cohort" or the other way round, to get a different view on your data. 

#### Metric variables

```{r density matrix}
dist_cat_breaks <- c(1,500,1000,2000,6000,100000)
dist_cat_labels <- c("< 500 km","500 - 1,000 km", "1,000 - 2,000 km", "2,000 - 6,000 km", "> 6,000 km")

age_groups    <- list(c(80,89),c(70,79),c(60,69),c(50,59),c(40,49),c(30,39),c(20,29))
period_groups <- list(c(1971,1979),c(1980,1989),c(1990,1999),c(2000,2009),c(2010,2018))
cohort_groups <- list(c(1980,1989),c(1970,1979),c(1960,1969),c(1950,1959),c(1940,1949),c(1930,1939),c(1920,1929))

plot_densityMatrix(dat              = travel,
                   y_var            = "mainTrip_distance",
                   age_groups       = age_groups,
                   period_groups    = period_groups,
                   log_scale        = TRUE)

plot_densityMatrix(dat                 = travel,
                   y_var               = "mainTrip_distance",
                   age_groups          = age_groups,
                   period_groups       = period_groups,
                   highlight_diagonals = list("born 1950 - 1959" = 8,
                                              "born 1970 - 1979" = 10),
                   log_scale           = TRUE)

plot_densityMatrix(dat              = travel,
                   y_var            = "mainTrip_distance",
                   age_groups       = age_groups,
                   period_groups    = period_groups,
                   log_scale        = TRUE,
                   y_var_cat_breaks = dist_cat_breaks,
                   y_var_cat_labels = dist_cat_labels,
                   highlight_diagonals = list("born 1950 - 1959" = 8,
                                              "born 1970 - 1979" = 10),
                   legend_title     = "Distance category")
```

If you change the plot type to "boxplot" you will get boxplots instead of densities.
```{r density matrix boxplot, warning = FALSE}
plot_densityMatrix(dat           = travel,
                   y_var         = "mainTrip_distance",
                   plot_type     = "boxplot",
                   age_groups    = age_groups,
                   period_groups = period_groups,
                   log_scale     = TRUE,
                   highlight_diagonals = list("born 1950 - 1959" = 8,
                                              "born 1970 - 1979" = 10))
```


#### Categorical variables

It is also possible to have a categorical variable as your y variable. Instead of densities, the absolute values of the
categories will be shown.
```{r density matrix categorical}
plot_densityMatrix(dat                 = travel,
                   y_var               = "household_size",
                   age_groups          = age_groups,
                   period_groups       = period_groups,
                   highlight_diagonals = list("born 1950 - 1959" = 8,
                                              "born 1970 - 1979" = 10))
```


# Model-based analyses

### Wrapper function for the estimation of mgcv::gam
To get an estimation for a further analysis, a wrapper function (just a function which wraps another function, in this case it sets the previous defined values for the age/period/cohort) is used to fit a generalized additive model (GAM). 
```{r}
model_pure <- gam(mainTrip_distance ~ te(age, period, bs = "ps", k = c(8,8)),
                  data = travel)
model_cov  <- gam(mainTrip_distance ~ te(age, period, bs = "ps", k = c(8,8)) +
                    residence_region + household_size + s(household_income),
                  data = travel)
```

### Partial effect plots
A partial APC plot highlights the interrelations between the temporal dimensions. These plots show the estimated odds ratios for the cohort effect dependent on age group and period. The mean marginal effect is marked as bold blue line. In addition to the marginal cohort effect, it includes one line for each partial cohort effect, that is, for the estimated differences between cohorts when just focusing on a specific age (left panel) or a specific period (right panel).

These plots show the estimated marginal odds ratios of age, period, and cohort for each distance category on a log2 scale. 
The dashed vertical lines in the cohort plot mark the boundaries between the generations defined in “Database” section.
```{r}
plot_partialAPCeffects(model_pure, dat = travel, variable = "period")
plot_marginalAPCeffects(model_pure, dat = travel, variable = "cohort",
                        vlines_vec = c(1939,1946,1966,1982,1994))

model_list <- list("pure model"      = model_pure,
                   "covariate model" = model_cov)
plot_jointMarginalAPCeffects(model_list, dat = travel,
                             vlines_list = list("cohort" = c(1939,1946,1966,1982,1994)))
```

### Non-smoothed heatmaps
Heatmaps show relationships between variables which are plotted on both axes. The make it easy to look for patterns in the cells by noticing the color change. The results are visualized by the estimated tensor product surface allowing the comparison of areas with a higher chance in the examined category and areas with a respective lower chance.
The Image starts with the observed heatmap, with binning the picture gets more distinct. After that, you choose which A, P, C dimension should be plotted on which axis and if reference lines should be included.

TODO binning is automatically performed, but can be turned off with the `bin_heatmap` argument.
When binning is used, the underlying grid used for binning can be set with the argument
`bin_heatmapGrid_list`.

```{r}
plot_APCheatmap(dat = travel, y_var = "mainTrip_distance", y_var_logScale = TRUE)
```

TODO APC axes can be set flexibly with the `dimensions` argument.

```{r}
plot_APCheatmap(dat = travel, y_var = "mainTrip_distance", dimensions = c("cohort","age"),
                y_var_logScale = TRUE)
```

### Smoothed, model-based heatmaps
This approach is used to get a heatmap of the model which was estimated prior. The tool will give you a smoothed heatmap with the possibility to and the lower/upper 95% CI boundary for the considered category.
```{r}
# estimated heatmap
APCtools::plot_APCheatmap(dat = travel, model = model_pure, plot_CI = FALSE,
                          bin_heatmap = FALSE,
                          markLines_list = list(cohort = c(1939,1946,1966,1982,1994)))

# binned heatmap
APCtools::plot_APCheatmap(dat = travel, model = model_pure, plot_CI = FALSE)

# including confidence intervals
APCtools::plot_APCheatmap(dat = travel, model = model_pure)
```

### Hexamaps
A commonly used display of APC data is the traditional Lexis diagram, which consists of a simple two-dimensional heatmap with a field of colored square tiles representing a quantity of interest. While it is informative, the main issue with a Lexis diagram is that the researcher has to make a choice to represent only two of the three dimensions (often age and period) on the XY axes, leaving cohorts to be represented on the diagonal. This setup introduces substantial visual distortion in how cohort patterns are presented relative to age and period.
In addition to these distortions, tracing cohort patterns is further complicated because adjacent square pixels along any cohort 
isoline only share a single corner. Together, these distortions can substantially impede cohort pattern recognition in Lexis diagrams.

A hexagonal grid is especially powerful for visualizing APC data because it places all three APC axes at equal 60° angles. Because of 
this placement, a hexamap overcomes all the visual distortions in a traditional Lexis diagram. So this tool helps to overcome many of the challenges of the existing APC data visualization tools.
```{r}
# restrict to data where the mortality rate is available
drug_deaths <- drug_deaths %>% filter(!is.na(mortality_rate))

# hexamap of an observed structure
plot_APChexamap(dat         = drug_deaths,
                y_var       = "mortality_rate",
                color_range = c(0,40))

# hexamap of a smoothed structure
model <- gam(mortality_rate ~ te(age, period, bs = "ps", k = c(8,8)),
             data = drug_deaths)

plot_APChexamap(dat = drug_deaths, model = model)
```

### Overview table

```{r}
create_APCsummary(model_list, dat = travel)
```

### Effect plots of linear and nonlinear effects
To show the effect of a single variable, this package offers effect plots of linear and nonlinear effects by using the estimated odds 
ratios of the variable in the model (for a specific category category) and presents the results on a log2 scale.Uncertainty is displayed by 95% confidence intervals. 
```{r}
APCtools::plot_linearEffects(model_cov)
APCtools::plot_1Dsmooth(model_cov, select = 2)
```

### Overview tables of linear and nonlinear effects

```{r}
summary_list <- create_covariateSummary(model_list)
summary_list[[1]]
summary_list[[2]]
```
